Improved state-space models for inference about spatial and temporal variation in abundance from count data

### 2013-03-29 Dail_Mad_ext.tex
Formatting and further trims

\documentclass[12pt]{article}

\usepackage[a4paper,hmargin={2.4cm,2.4cm},vmargin={2.4cm,2.4cm}]{geometry}
\usepackage[mathlines,displaymath]{lineno}
\usepackage{setspace}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{bm}
\usepackage{graphicx}
\usepackage[round,colon,authoryear]{natbib}
\usepackage{color, soul}
\usepackage{fancyvrb}
\usepackage{verbatim}
\usepackage[pdftex,colorlinks=true,citecolor=black,linkcolor=black,urlcolor=black,pdftex,pdfstartview=FitH,bookmarks=true]{hyperref}
\usepackage{rotating}

\bibliographystyle{ecology}
\hyphenpenalty=100000

@@

\abstract{
Models of population dynamics play a central role in theoretical and
applied ecology where they are used for purposes such as testing 
hypotheses about density dependence and predicting species' responses 
to future environmental change or conservation actions. Failure to
account for observation error in such models can result in 
bias, and thus ecologists have increasingly relied on state-space
models to directly model both the observation error and the ecological
process of interest.
Conventional state-space models, however, have
three important limitations: (1) the parameters are not identifiable
in many common situations, (2) they do not admit spatial variation in
population dynamics, and (3) there is no clear interpretation of the
observation error. We demonstrate how each of these problems can be
resolved using a class of hierarchical models proposed by
\citet[Biometrics]{dail_madsen:2011} for spatially-replicated
time-series data that attributes observation error to imperfect detection.
We expand this class of models to accommodate classical growth
models (e.g. exponential and Ricker-logistic), zero-inflation,
and random effects such as observer-specific
detection probabilities. We also present methods for forecasting
population size under future environmental conditions. Implementation
of these ideas is possible using either frequentist or Bayesian
methods, and code to fit these models using the \textbf{R} package
\texttt{unmarked} and \textbf{JAGS} is also provided.
A simulation study indicated that bias was negligible and coverage
nominal for the proposed model extensions. An analysis of data from
the North American Breeding Bird Survey highlighted how these methods
can be readily applied to existing data, but it also suggested the
that precision will be low when direct information about detection
probability (such as is collected using distance sampling or
replicated counts) is unavailable.
}

@@

Bayesian inference has several appealing features. First, it
allows direct probability statements to be made about a hypothesis
given data \citep{link_barker:2010}.    
Second, Bayesian methods offer straight-forward approaches for
combine data from multiple sources or use existing estimates of parameters as prior
distributions. In practice, simulation methods such as Markov
chain Monte Carlo (MCMC) are used to simulate the posterior
distributions. %Although MCMC can be slow, our experience is that it
%may actually converge faster than the amount of time required to
%maximize the likelihood shown in Eq.\ref{eq:like}, especially
%when the upper bound of the summation must be high, e.g. $N_{max}>200$.
%Furthermore, 
MCMC may be the only possible option for estimating parameters
using some of the extensions described below, such as allowing for
additional random effects.

@@

Although partitioning population growth into survival and recruitment
is ideal in terms of providing a mechanistic description of population
dynamics, it is not always possible using simple count data,
especially when the sites are not closed with respect to immigration
and emigration. %Furthermore, a
%simpler model would have several merits: faster running time, fewer
%total model combinations, and possibly more realistic estimates.
If the mechanistic model must be sacrificed, we suggest replacing it
with classical population growth models that have a long history in
ecology. The simplest of which is the exponential growth
model: $f(N_{it-1}) = N_{it-1}e^r$ where $r$ is the intrinsic
rate of increase. Although demographic and environmental stochasticity
in $r$ could be modeled as functions of covariates or random effects, a
simpler way of allow for stochasticity is to regard $N_{it}$ as a Poisson random variable, simplifying
Eq.~\ref{eq:Nt} to:
\begin{equation}
  N_{it} \sim \text{Pois}(\exp(r)N_{it-1}).
\label{eq:exp}
\end{equation}
This,
like the geometric-recruitment version of the model, is a variant on a simple
density-independent exponential model of population growth.
Density-dependent versions of the model are also possible.  For
example:
\begin{equation}
  N_{it} \sim \text{Pois}(N_{it-1}\exp(r(1-N_{it-1}/K)))
\label{eq:rick}
\end{equation}
where $K$ is the stable equilibrium of the population and $r$ is
the
instantaneous population growth rate at low population
densities, and
both parameters are constrained to be positive. This is a
stochastic version
of the Ricker-logistic population growth model
\citep{ricker:1954}. We also
implemented a modified Gompertz-logistic density-dependent
model \citep{hart_gotelli:2011}:
\begin{equation}
N_{it} \sim
\text{Pois}(N_{it-1}\exp(r(1-\log(N_{it-1}+1)/\log(K+1))))
\label{eq:gomp}
\end{equation}
Here the interpretations of $r$ and $K$ are similar to in the
Ricker-logistic model.

@@

The geometric-recruitment, exponential, Ricker-logistic, and Gompertz-logistic
versions of the DM models all share a common feature (or bug):
once the population at a site reaches 0, it must remain at 0.
This is because all contributions to population growth are
local in these models.  We generalized each these models that
include both internal and external (immigration) contributions
to population growth.  The exponential plus immigration
model is:
\begin{equation}
  N_{it} \sim \text{Poisson}(\exp(r)N_{it-1} + \iota)
  \label{eq:expimm2}
\end{equation}
where $\iota$ represents the immigration rate, and is
constrained to be
positive (this is equivalent to separate Poisson processes for
growth and immigration).  This model is close to
the constant-recruitment DM model (Eq.~\ref{eq:Nt}), with $\exp(r)$ instead
of $\omega$ and $\iota$ instead of $\gamma$, except that the first
process is Poisson distributed instead of binomial. The geometric-recruitment, Ricker-logistic,
and Gompertz-logistic models can be extended to allow for immigration in the
same way.   Deterministic versions of the Ricker-logistic + immigration and the 
Gompertz-logistic + immigration do reach stable equilibriums, but the equilibriums are not at $K$ and cannot be solved for analytically \citep{otto_day:2007}.  Therefore, we
refer to $K$ in these models as the semi-equilibrium abundance.
\begin{comment}
Similarly, the Ricker-logistic + immigration model can be represented as:
the
Gompertz-logistic + immigration model as: (13) and the geometric-recruitment +  immigration model as: (14)
\end{comment}

@@

Variation in detection probability among observers is another
source of random variation that may need to be accounted for applying
the DM model. For example, differences in observers' ability to see,
hear, or identify birds has long been recognized as a potential source of error
in avian point count surveys such as the BBS 
\citep{robbins_etal:1986,sauer_etal:1994auk,diefenbach_etal:2003,campbell_francis:2011}.%,alldredge_etal:2007auk
%Estimating a separate detection probability for each observer
%can be difficult and reduces one's ability to estimate the quantities
%of interest.  This problem is compounded by the fact that
%observers differ greatly in the number of surveys they have
%run (so that many observers' separate detection probabilities
%could not be accurately estimated).

@@

%GENERAL ABOUT BBS DATA.
The BBS is an annual roadside survey implemented by trained
observers in the United States and Canada. An observer conducts 50
3-minute point counts, with 400 m radiuses and 0.8 km apart from each other along a
39.4 km route. We only used data marked as acceptable for use in the annual BBS
analysis \citep{sauer_etal:1994auk}.  We summed the total birds of each
species seen on each route and year and used the routes (rather
than the individual stops) as our sites.
Strong winds can interfere with point count observers' ability
to hear birds \citep{simons_etal:2007}; we tested the effects of wind
speed on detection probability. BBS volunteers record wind conditions
at the beginning and end of each route on the Beaufort Scale
\citep[start and end wind]{robbins_etal:1986}.
%(start and end wind; Robbins et al. 1986)(start and end wind;0-9; cite).
When start or end wind was not recorded we imputed those values with
the start or end mean. We took the mean of start and end wind for
each route and put this average wind scale value into four
categories: 0 $\leq$ wind $<$ 1; 1 $\geq$ wind $<$ 2; 2 $\leq$ wind $<$ 3; and wind
$\geq$ 3 (maximum of 3.5).
Following \citet{link_sauer:2002},
%Link and Sauer (2002),
we also included the first time an observer ran a route as a predictor variable for detection
probability.

%HOW WE OBTAINED AND MANIPULATED PRISM DATA
We ran a series of maximum-likelihood based models, 
and then a series of Bayesian models. We started by testing
three models of initial abundance (Poisson, negative binomial, and
zero-inflated Poisson) with exponential dynamics
(Eq.~\ref{eq:exp}) and no covariates.  We selected the minimum Akaike's Information
Criterion (AIC) model from that set to test three additional
models for p: wind, first, and wind + first. We selected the minimum
AIC model from that set to test eight additional models of dynamics:
constant-recruitment, geometric-recruitment, Ricker-logistic, Gompertz-logistic, geometric-recruitment +
immigration, exponential + immigration, Ricker-logistic +
immigration, and Gompertz-logistic + immigration. %We then tested the effect of average
%minimum temperature for June and July on all dynamics parameters from
%the minimum AIC dynamics model.
As above, these models require a a finite upper bound ($N_{max}$) to integrate
over when run in a maximum likelihood
framework; we used 600. % for ovenbirds and 350 for golden-winged warblers.

We ran the top ranked models from the maximum likelihood
analyses in a Bayesian framework.  We added random observer 
effects and environmental stochasticity (modeled as lognormal
variation in regional population growth rate, Eq.~\ref{eq:nuRand}). % and, where appropriate, zero-inflated dynamics.  
We used
non-informative priors. We tested for lack of convergence using
5 Markov chains for each model \citep{gelman_rubin:1992}.
%(Gelman and Rubin 1992).
For each chain we sampled the MCMC for at least 40,000 iterations, after at
least 3,000 adaptation and tuning samples.

We also used these models to obtain route and 
year specific abundance estimates.  Because of the memory requirements involved,
we obtained these estimates from a single chain.  %We present these estimates for ovenbirds only.
%Latent inclusion factors for temperature effects?

@@

The negative binomial distribution was strongly supported for ovenbird
initial abundance over the Poisson and zero-inflated Poisson
(Table 2A). %Even compared to the Poisson, there was
%no evidence to support a zero-inflation factor for this species.
The best supported model for $p$ was additive effects of wind speed
and first run (Table 2B, model B.1). First run and increasing wind
speeds both decreased $p$. All dynamics models with immigration were
better supported than models without immigration (Table 2C); 
the best supported of these was the Ricker-logistic + immigration.
%The best supported model of those without immigration was the
%Gompertz-logistic, and the geometric-recruitment and the constant model had the least
%support.
%There was little support for an effect of minimum June and July
% temperature on dynamics parameters of the Ricker + Immigration model
%(Table 2A, models A.17 - A.18).

Estimates for the top ranked model for ovenbird
%(NB[$\Lambda (.) \alpha (.)$]Ricker-logistic+Immigration[$r(.)K(.) \iota(.)$]p(wind+1st)) 
were similar when
run in the Bayesian framework, except that the estimate of $r$
more than halved.
% (from 0.026 $\pm$ 0.006 to 0.011 $\pm$ 0.005 [for
%Bayesian model estimates we present mean and SD]). 
When random observer effects
were added, estimates for $\Lambda$ and $K$ increased dramatically,
%(from 31.6 $\pm$ 4.9 to 42.5 $\pm$ 7.3 and from 56.2 $\pm$ 18.2 to 101.8 $\pm$ 21.1,
%respectively)
and the estimate for r was intermediate. The %(0.018 $\pm$ 0.006)
estimate of the intercept for $p$ (on the logit scale) dropped. % from -1.5 $\pm$
%0.1 to -2.0 $\pm$ 0.1, and the estimate of $\sigma_p$ was 0.35 $\pm$ 0.03.  
When environmental stochasticity was added, the estimate for $K$ increased. %(to 112 $\pm$ 58.8) and $r$ was
%again intermediate (0.020 $\pm$ 0.010).  Adding environmental stochasticity alone had little effect
%on estimate of $p$ or $\Lambda$. 
The highest estimates of $\Lambda$ and $K$ came 
from the model that included environmental stochasticity and random observer effects. %(43.8 $\pm$
%7.9 and 141.5 $\pm$ 69.5, respectively).  
Gelman and Rubin diagnostics and visual examination of the chain trajectory and density plots
provided some evidence for a lack of convergence in estimates of the
intercept for $p$ for all Bayesian models ran.

@@

%For both species, 
Constant-recruitment dynamics models estimated very high
survival probabilities ($\omega$ = 1 $\pm$ 1.3e-05). % for ovenbird and
%0.935 $\pm$ 0.011 for
%golden-winged warbler). 
The geometric-recruitment and geometric-recruitment + immigration
models estimated very low survival probabilities %for ovenbird 
($\omega$ =
0.058 $\pm$ 0.046 and 0.026 $\pm$ 0.060, respectively). 
%but not for golden-winged warbler ($\omega$ = 0.715 $\pm$ 0.288 for both models).

@@

Our work has highlighted the limitations of classical state-space
models as applied to ecological time series data, and we have
demonstrated how a class of open population $N$-mixture models
proposed by \citet{dail_madsen:2011} overcomes these
limitations. We extended their model in several important
ways to accomodate common features of ecological
datasets, including sparse counts, zero-inflation, environmental
stochasticity, and random variation in observation error.
%We made four important developments of the open population
%N-mixture model proposed by \citet{dail_madsen:2011} that make the
%model more applicable to long-term data such as is collected
%by the North American Breeding Bird Survey.
%First, we have
%reconciled the objectives of traditional state-space models
%with open population $N$-mixture models by illustrating how
%classical models of population dynamics can be embedded within
%the framework. Second, we have demonstrated methods of
%accounting for zero-inflation in the time-series. Third, we
%have illustrated how additional random effects such as
%observer-specific detection probabilities can be
%accommodated. Fourth, we presented a Bayesian analysis of the
%model, which, makes it possible to incorporate prior information when
%available, and in some cases, facilitates parameter estimation.
We also demonstrated how many of the objectives of conventional
state-space modeling can be accomplished within this expanded
framework. For example, one of the primary aims of our paper was to demonstrate how classical
population growth models can be embedded in the DM model.
Although we view this as an important connection to make between the two
different classes of state-space models, we believe that
%In some ways,
such population growth models are more phenomenological
than mechanistic, and this runs counter to the motivation for
the DM model. However, estimating
demographic parameters from count data is an ambitious goal, and
the required assumption will not be valid in many cases,
especially when immigration and emigration occur. This was supported by the
unrealistic estimates of vital rates reported here and by the
lack of support for the mechanistic models relative to the population
growth models.

@@

Although these integrated population models may be viewed as the gold standard in
state-space modeling, ecologists are not always so fortunate to have direct
information about vital rates, especially at large spatial scales.
Rather, count data are much more common, and
are produced by many of the largest monitoring programs in the
world, such as the North American Breeding Bird Survey
\citep[BBS;][]{robbins_etal:1986}.
In the absence of direct information about demographic parameters, and
when the original assumptions of the model do not hold, the
model can still be used for important purposes such as predicting future
population size and changes in species distributions.

@@

We have presented a test case for these models using data from
the BBS for ovenbirds. We demonstrate that these models can be
used to model and estimate initial abundance, detection
probability, and density-independent and dependent functions for
population dynamics. We tested these models in both the
frequentist and Bayesian frameworks and found strengths and
weaknesses of each approach. The Bayesian analyses permit more
flexibility and allow the incorporation of random observer
effects, environmental stochasticity, and zero-inflated dynamics. 
Our results suggest ignoring
variation between observers can lead to underestimation of
initial abundance and equilibrium population sizes. On the other
hand, our Bayesian analyses did have some convergence problems.
%particularly for the species (golden-winged warbler) with sparser data. 
It's likely that poor model fits are related to the low estimates of
detection probability, which in turn are probably caused by the
large 400 m radius used for BBS point counts. Again, we believe
that direct information on detectability (such as repeated
visits within a season) would greatly improve the fit of these
models in both frameworks.  We also were able to obtain year
and site-specific estimates of abundance (Fig.~\ref{fig:oven_N}),
although with generally lower precision than more global estimates.  These sort of estimates
may be more reliable when abundance and dynamics are modeled 
as functions of habitat covariates, which we did not do here.
\begin{comment}
[Alternatively, could put discussions of our results where
they fit in by subject area, rather than separately. For
instance, could discuss BBS population dynamics model selection
in population models paragraph, BBS Bayesian vs.
frequentist strengths and weakness in the Bayesian paragraph, 
and golden-winged warbler ZIP results in ZIP
paragraph.]
\end{comment}

>>>>

\documentclass[12pt]{article}
\renewcommand*\abstractname{Summary}
\usepackage[a4paper,hmargin={2.4cm,2.4cm},vmargin={2.4cm,2.4cm}]{geometry}
\usepackage[mathlines,displaymath]{lineno}
\usepackage{setspace}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{bm}
\usepackage{graphicx}
\usepackage[round,colon,authoryear]{natbib}
\usepackage{color, soul}
\usepackage{fancyvrb}
\usepackage{verbatim}
\usepackage[pdftex,colorlinks=true,citecolor=black,linkcolor=black,urlcolor=black,pdftex,pdfstartview=FitH,bookmarks=true]{hyperref}
\usepackage{rotating}

\bibliographystyle{ecology}
%\hyphenpenalty=100000

@@

\abstract{
\noindent
\begin{enumerate}
\item Models of population dynamics play a central role in theoretical and
applied ecology, where they are used for purposes such as testing 
hypotheses about density dependence and predicting species' responses 
to future environmental change or conservation actions. Failure to
account for observation error in such models can result in 
bias, and thus ecologists have increasingly relied on state-space
models to directly model both the observation error and the ecological
process of interest.
\item Conventional state-space models, however, have
three important limitations: (1) the parameters are not identifiable
in many common situations, (2) they do not admit spatial variation in
population dynamics, and (3) there is no clear interpretation of the
observation error. We demonstrate how each of these problems can be
resolved using a class of hierarchical models proposed by
\citet[Biometrics]{dail_madsen:2011} for spatially-replicated
time-series data that attributes observation error to imperfect detection.
\item We expand this class of models to accommodate classical growth
models (e.g. exponential and Ricker-logistic), zero-inflation,
and random effects such as observer-specific
detection probabilities. We also present methods for forecasting
population size under future environmental conditions. Implementation
of these ideas is possible using either frequentist or Bayesian
methods, and code to fit these models using the \textbf{R} package
\texttt{unmarked} and \textbf{JAGS} is also provided.
\item A simulation study indicated that bias was negligible and coverage
nominal for the proposed model extensions. An analysis of data from
the North American Breeding Bird Survey highlighted how these methods
can be readily applied to existing data, but it also suggested the
that precision will be low when direct information about detection
probability (such as is collected using distance sampling or
replicated counts) is unavailable.
\end{enumerate}
}

@@

However, Bayesian inference has several appealing features. First, it
allows direct probability statements to be made about a hypothesis
given data \citep{link_barker:2010}.    
Second, Bayesian methods offer straight-forward approaches for
combine data from multiple sources or use existing estimates of parameters as prior
distributions. In practice, simulation methods such as Markov
chain Monte Carlo (MCMC) are used to simulate the posterior
distributions. %Although MCMC can be slow, our experience is that it
%may actually converge faster than the amount of time required to
%maximize the likelihood shown in Eq.\ref{eq:like}, especially
%when the upper bound of the summation must be high, e.g. $N_{max}>200$.
%Furthermore, 
MCMC may be the only possible option for estimating parameters
using some of the extensions described below, such as allowing for
additional random effects.

@@

Although partitioning population growth into survival and recruitment
is ideal in terms of providing a mechanistic description of population
dynamics, it is not always possible using simple count data,
especially when the sites are not closed with respect to immigration
and emigration. %Furthermore, a
%simpler model would have several merits: faster running time, fewer
%total model combinations, and possibly more realistic estimates.
If the mechanistic model must be sacrificed, we suggest replacing it
with classical population growth models that have a long history in
ecology. The simplest of which is the exponential growth
model: $f(N_{it-1}) = N_{it-1}e^r$ where $r$ is the intrinsic
rate of increase. %Although demographic and environmental stochasticity
%in $r$ could be modeled as functions of covariates or random effects, a
%simpler way of 
To allow for demographic stochasticity, we can regard $N_{it}$ as a Poisson 
random variable, simplifying Eq.~\ref{eq:Nt} to:
\begin{equation}
  N_{it} \sim \text{Pois}(\exp(r)N_{it-1}).
\label{eq:exp}
\end{equation}
This,
like the geometric-recruitment version of the model, is a variant on a simple
density-independent exponential model of population growth.
Density-dependent versions of the model are also possible.  For
example:
\begin{equation}
  N_{it} \sim \text{Pois}(N_{it-1}\exp(r(1-N_{it-1}/K)))
\label{eq:rick}
\end{equation}
where $K$ is the stable equilibrium of the population and $r$ is
the
instantaneous population growth rate at low population
densities, and
both parameters are constrained to be positive. This is a
stochastic version
of the Ricker-logistic population growth model
\citep{ricker:1954}. We also
implemented a modified Gompertz-logistic density-dependent
model \citep{hart_gotelli:2011}:
\begin{equation}
N_{it} \sim
\text{Pois}(N_{it-1}\exp(r(1-\log(N_{it-1}+1)/\log(K+1))))
\label{eq:gomp}
\end{equation}
Here the interpretations of $r$ and $K$ are similar to in the
Ricker-logistic model.

@@

The geometric-recruitment, exponential, Ricker-logistic, and Gompertz-logistic
versions of the DM models all share a common feature (or bug):
once the population at a site reaches 0, it must remain at 0.
This is because all contributions to population growth are
local in these models.  We generalized each these models that
include both internal and external (immigration) contributions
to population growth.  The exponential plus immigration
model is:
\begin{equation}
  N_{it} \sim \text{Poisson}(\exp(r)N_{it-1} + \iota)
  \label{eq:expimm2}
\end{equation}
where $\iota$ represents the immigration rate, and is
constrained to be
positive (this is equivalent to separate Poisson processes for
growth and immigration).  %This model is close to
%the constant-recruitment DM model (Eq.~\ref{eq:Nt}), with $\exp(r)$ instead
%of $\omega$ and $\iota$ instead of $\gamma$, except that the first
%process is Poisson distributed instead of binomial. 
The geometric-recruitment, Ricker-logistic,
and Gompertz-logistic models can be extended to allow for immigration in the
same way.   Deterministic versions of the Ricker-logistic + immigration and the 
Gompertz-logistic + immigration do reach stable equilibriums, but the equilibriums 
are not at $K$ and cannot be solved for analytically \citep{otto_day:2007}.  Therefore, we
refer to $K$ in these models as the semi-equilibrium abundance.
\begin{comment}
Similarly, the Ricker-logistic + immigration model can be represented as:
the
Gompertz-logistic + immigration model as: (13) and the geometric-recruitment +  immigration model as: (14)
\end{comment}

@@

Variation in detection probability among observers is another
source of random variation that may need to be accounted for applying
the DM model. For example, differences in observers' ability to see,
hear, or identify birds has long been recognized as a potential source of error
in avian point count surveys such as the BBS 
\citep{robbins_etal:1986,sauer_etal:1994auk,campbell_francis:2011}.%,alldredge_etal:2007auk %diefenbach_etal:2003,
%Estimating a separate detection probability for each observer
%can be difficult and reduces one's ability to estimate the quantities
%of interest.  This problem is compounded by the fact that
%observers differ greatly in the number of surveys they have
%run (so that many observers' separate detection probabilities
%could not be accurately estimated).

@@

%GENERAL ABOUT BBS DATA.
The BBS is an annual roadside survey implemented by trained
observers in the United States and Canada. An observer conducts 50
3-minute point counts, with 400 m radiuses and 0.8 km apart from each other along a
39.4 km route. We only used data marked as acceptable for use in the annual BBS
analysis \citep{sauer_etal:1994auk}.  We summed the total birds of each
species seen on each route and year and used the routes (rather
than the individual stops) as our sites.
Strong winds can interfere with point count observers' ability
to hear birds \citep{simons_etal:2007}; we tested the effects of wind
speed on detection probability, using the truncated mean of wind conditions at the start and end of
each route on the Beaufort scale as a categorical predictor. 
%BBS volunteers record wind conditions
%at the beginning and end of each route on the Beaufort Scale
%\citep[start and end wind]{robbins_etal:1986}.
%%(start and end wind; Robbins et al. 1986)(start and end wind;0-9; cite).
%When start or end wind was not recorded we imputed those values with
%the start or end mean. We took the mean of start and end wind for
%each route and put this average wind scale value into four
%categories: 0 $\leq$ wind $<$ 1; 1 $\geq$ wind $<$ 2; 2 $\leq$ wind $<$ 3; and wind
%$\geq$ 3 (maximum of 3.5).
Following \citet{link_sauer:2002},
%Link and Sauer (2002),
we also included the first time an observer ran a route as a predictor variable for detection
probability.

We ran a series of frequentist models, 
and then a series of Bayesian models. We started by testing
three models of initial abundance (Poisson, negative binomial, and
zero-inflated Poisson) with exponential dynamics
(Eq.~\ref{eq:exp}) and no covariates.  We selected the minimum Akaike's Information
Criterion (AIC) model from that set to test three additional
models for p: wind, first, and wind + first. We selected the minimum
AIC model from that set to test eight additional models of dynamics:
constant-recruitment, geometric-recruitment, Ricker-logistic, Gompertz-logistic, 
the previous three with immigration, and exponential + immigration.
%geometric-recruitment +
%immigration, exponential + immigration, Ricker-logistic +
%immigration, and Gompertz-logistic + immigration. %We then tested the effect of average
%minimum temperature for June and July on all dynamics parameters from
%the minimum AIC dynamics model.
As above, these models require a a finite upper bound ($N_{max}$) to integrate
over when run in a maximum likelihood
framework; we used 600. % for ovenbirds and 350 for golden-winged warblers.

We ran the top ranked models from the maximum likelihood
analyses in a Bayesian framework.  We added random observer 
effects and environmental stochasticity (modeled as lognormal
variation in regional population growth rate, Eq.~\ref{eq:nuRand}). % and, where appropriate, zero-inflated dynamics.  
We used
non-informative priors. We tested for lack of convergence using
5 Markov chains for each model \citep{gelman_rubin:1992}.
%(Gelman and Rubin 1992).
For each chain we sampled the MCMC for at least 40,000 iterations, after at
least 3,000 adaptation and tuning samples.  
We also used these models to obtain route and 
year specific abundance estimates.  Because of the memory requirements involved,
we obtained these estimates from a single chain.  %We present these estimates for ovenbirds only.
%Latent inclusion factors for temperature effects?

@@

The negative binomial distribution was strongly supported for ovenbird
initial abundance over the Poisson and zero-inflated Poisson
(Table 2A). %Even compared to the Poisson, there was
%no evidence to support a zero-inflation factor for this species.
The best supported model for $p$ was additive effects of wind speed
and first run (Table 2B, model B.1). First run and increasing wind
speeds both decreased $p$. All dynamics models with immigration were
better supported than models without immigration (Table 2C); 
the best supported of these was the Ricker-logistic + immigration.
%The best supported model of those without immigration was the
%Gompertz-logistic, and the geometric-recruitment and the constant model had the least
%support.
%There was little support for an effect of minimum June and July
% temperature on dynamics parameters of the Ricker + Immigration model
%(Table 2A, models A.17 - A.18).
%For both species, 
Constant-recruitment dynamics models estimated very high
survival probabilities ($\omega$ = 1 $\pm$ 1.3e-05). % for ovenbird and
%0.935 $\pm$ 0.011 for
%golden-winged warbler). 
The geometric-recruitment and geometric-recruitment + immigration
models estimated very low survival probabilities %for ovenbird 
($\omega$ =
0.058 $\pm$ 0.046 and 0.026 $\pm$ 0.060, respectively). 
%but not for golden-winged warbler ($\omega$ = 0.715 $\pm$ 0.288 for both models).

Estimates for the top ranked model %for ovenbird
%(NB[$\Lambda (.) \alpha (.)$]Ricker-logistic+Immigration[$r(.)K(.) \iota(.)$]p(wind+1st)) 
were similar when
run in the Bayesian framework, except that the estimate of $r$
more than halved.
% (from 0.026 $\pm$ 0.006 to 0.011 $\pm$ 0.005 [for
%Bayesian model estimates we present mean and SD]). 
When random observer effects
were added, estimates for $\Lambda$ and $K$ increased dramatically,
%(from 31.6 $\pm$ 4.9 to 42.5 $\pm$ 7.3 and from 56.2 $\pm$ 18.2 to 101.8 $\pm$ 21.1,
%respectively)
and the estimate for r was intermediate. The %(0.018 $\pm$ 0.006)
estimate of the intercept for $p$ (on the logit scale) dropped. % from -1.5 $\pm$
%0.1 to -2.0 $\pm$ 0.1, and the estimate of $\sigma_p$ was 0.35 $\pm$ 0.03.  
When environmental stochasticity was added, the estimate for $K$ increased. %(to 112 $\pm$ 58.8) and $r$ was
%again intermediate (0.020 $\pm$ 0.010).  Adding environmental stochasticity alone had little effect
%on estimate of $p$ or $\Lambda$. 
The highest estimates of $\Lambda$ and $K$ came 
from the model that included environmental stochasticity and random observer effects. %(43.8 $\pm$
%7.9 and 141.5 $\pm$ 69.5, respectively).  
Gelman and Rubin diagnostics and visual examination of the chain trajectory and density plots
provided some evidence for a lack of convergence in estimates of the
intercept for $p$ for all Bayesian models ran.

@@

@@

Our work has highlighted the limitations of classical state-space
models as applied to ecological time series data, and we have
demonstrated how a class of open population $N$-mixture models
proposed by \citet{dail_madsen:2011} overcomes these
limitations. We extended their model in several important
ways to accomodate common features of ecological
datasets, including sparse counts, zero-inflation, environmental
stochasticity, and random variation in observation error.
%We made four important developments of the open population
%N-mixture model proposed by \citet{dail_madsen:2011} that make the
%model more applicable to long-term data such as is collected
%by the North American Breeding Bird Survey.
%First, we have
%reconciled the objectives of traditional state-space models
%with open population $N$-mixture models by illustrating how
%classical models of population dynamics can be embedded within
%the framework. Second, we have demonstrated methods of
%accounting for zero-inflation in the time-series. Third, we
%have illustrated how additional random effects such as
%observer-specific detection probabilities can be
%accommodated. Fourth, we presented a Bayesian analysis of the
%model, which, makes it possible to incorporate prior information when
%available, and in some cases, facilitates parameter estimation.
We also demonstrated how many of the objectives of conventional
state-space modeling can be accomplished within this expanded
framework. For example, one of the primary aims of our paper was to demonstrate how classical
population growth models can be embedded in the DM model.
Although we view this as an important connection to make between the two
different classes of state-space models, we believe that
%In some ways,
such population growth models are more phenomenological
than mechanistic, and this runs counter to the motivation for
the DM model. However, estimating
demographic parameters from count data is an ambitious goal, and
the required assumptions will not be valid in many cases,
especially when immigration and emigration occur. This was supported by the
unrealistic estimates of vital rates reported here and by the
lack of support for the mechanistic models relative to the population
growth models.

@@

Although these integrated population models may be viewed as the gold standard in
state-space modeling, ecologists are not always so fortunate to have direct
information about vital rates, especially at large spatial scales.
Rather, count data are much more common, and
are produced by many of the largest monitoring programs in the
world, such as the BBS. %North American Breeding Bird Survey
%\citep[BBS;][]{robbins_etal:1986}.
In the absence of direct information about demographic parameters, and
when the original assumptions of the model do not hold, the
model can still be used for important purposes such as predicting future
population size and changes in species distributions.

@@

%We have presented a test case for these models using data from
%the BBS for ovenbirds. We demonstrate that these models can be
%used to model and estimate initial abundance, detection
%probability, and density-independent and dependent functions for
%population dynamics. We tested these models in both the
%frequentist and Bayesian frameworks and found strengths and
%weaknesses of each approach. The Bayesian analyses permit more
%flexibility and allow the incorporation of random observer
%effects, environmental stochasticity, and zero-inflated dynamics. 
%Our results suggest ignoring
%variation between observers can lead to underestimation of
%initial abundance and equilibrium population sizes. On the other
%hand, our Bayesian analyses did have some convergence problems.
%%particularly for the species (golden-winged warbler) with sparser data. 
%It's likely that poor model fits are related to the low estimates of
%detection probability, which in turn are probably caused by the
%large 400 m radius used for BBS point counts. Again, we believe
%that direct information on detectability (such as repeated
%visits within a season) would greatly improve the fit of these
%models in both frameworks.  We also were able to obtain year
%and site-specific estimates of abundance (Fig.~\ref{fig:oven_N}),
%although with generally lower precision than more global estimates.  These sort of estimates
%may be more reliable when abundance and dynamics are modeled 
%as functions of habitat covariates, which we did not do here.
\begin{comment}
[Alternatively, could put discussions of our results where
they fit in by subject area, rather than separately. For
instance, could discuss BBS population dynamics model selection
in population models paragraph, BBS Bayesian vs.
frequentist strengths and weakness in the Bayesian paragraph, 
and golden-winged warbler ZIP results in ZIP
paragraph.]
\end{comment}
